High-level cognition is supported by at least second order dynamic correlations in neural activity patterns

### 2019-09-05 paper/main.tex
finished draft of discussion

\section*{Introduction}
A central goal in cognitive neuroscience is to elucidate the
\textit{neural code}: the mapping between (a) mental states or
cognitive representations and (b) neural activity patterns. One means
of testing models of the neural code is to ask how accurately that
model is able to ``translate'' neural activity patterns into known (or
hypothesized) mental states or cognitive
representations~\citep[e.g.,][]{HaxbEtal01, NormEtal06, TongPrat12,
  MitcEtal08a, KamiTong05, NishEtal11, PereEtal18, HuthEtal12,
  HuthEtal16}.  Training decoding models on different types of neural
features (Fig.~\ref{fig:patterns}a) can also help to elucidate which
specific aspects of neural activity patterns are informative about
cognition-- and, by extension, which types of neural activity patterns
might comprise the neural code.  For example, prior work has used
region of interest analyses to estimate the anatomical locations of
specific neural representations~\citep[e.g.,][]{EtzeEtal09}, or to
compare the relative contributions to the neural code of multivariate
activity patterns versus patterns of dynamic correlations between
neural activity patterns~\citep[e.g.,][]{MannEtal18, FongEtal19}.  An
emerging theme in this literature is that cognition is mediated by
complex dynamic interactions between brain
structures~\citep{SporHone06, BassEtal06, Turk13, DemeEtal19}.

@@

Studies of the neural code to date have primarily focused on
univariate or multivariate neural patterns~\citep[for review
see][]{NormEtal06}, or (more recently) on patterns of dynamic
first-order correlations~\citep[i.e., interactions between pairs of
brain structures;][]{MannEtal18, FongEtal19}.  We wondered what the
future of this line of work might hold.  For example, is the neural
code mediated by higher-order interactions between brain structures?
Second-order correlations reflect \textit{homologous} patterns of
correlation.  In other words, if the changing patterns of correlations
between two regions, $A$ and $B$, are similar to those between two
other regions, $C$ and $D$, this would be reflected in the
second-order correlations between ($A$--$B$) and ($C$--$D$).  In this
way, second-order correlations identify similarities and differences
between subgraphs of the brain's connectome.  Analogously, third-order
correlations reflect homologies between second-order correlations--
i.e., homologous patterns of homologous interactions between brain
regions.  More generally, higher-order correlations reflect homologies
between patterns of lower-order correlations.  We can then ask: which
``orders'' of interaction are most reflective of high-level cognitive
processes?

@@

To gain insights into the above aspects of the neural code, we
developed a computational framework for estimating dynamic high-order
correlations in timeseries data. This framework provides an important
advance, in that it enables us to examine patterns in higher-order
correlations that are computationally intractable to estimate via
conventional methods.  Given a multivariate timeseries, our framework
provides timepoint-by-timepoint estimates of the first-order
correlations, second-order correlations, and so on (up to tenth-order
correlations in this manuscript).  Our approach combines a
kernel-based method for computing dynamic correlations in timeseries
data with a dimensionality reduction step (Fig.~\ref{fig:patterns}b)
that projects the resulting dynamic correlations into a
low-dimensional space.  We explored two dimensionality reduction
approaches: principle components analysis~\citep[PCA;][]{Pear01},
which preserves an approximately invertable transformation back to the
original data; and a second non-invertible algorithm that explored
patterns in eigenvector centrality~\citep{Land95}.  This latter
approach characterizes correlations between each feature dimension's
relative \textit{position} in the network in favor of the specific
activity histories of different features.

@@

\subsection*{Kernel-based approach for computing dynamic correlations}
Given a matrix of observations, we can compute the (static)
Pearson's correlation between any pair of columns, $\mathbf{X}(\cdot, i)$ and
$\mathbf{X}(\cdot, j)$ using~\citep{Pear01}:
\begin{align}
  \mathrm{corr}(\mathbf{X}(\cdot, i), \mathbf{X}(\cdot, j)) &=
                                                              \frac{\sum_{t=1}^T
                                                              \left(\mathbf{X}(t,
                                                              i)
                                                              -
                                                              \bar{\mathbf{X}(\cdot,
                                                              i)}\right)
                                                              \left(\mathbf{X}(t,
                                                              j)
                                                              -
                                                              \bar{\mathbf{X}(\cdot, j)}\right)}{\sqrt{\sum_{t=1}^T
                                                              \sigma^2_{\mathbf{X}(\cdot, i)} 
                                                              \sigma^2_{\mathbf{X}(\cdot, j)}}},~\mathrm{where}\\\label{eqn:corr}
  \bar{\mathbf{X}(\cdot, k)} &= \sum_{t=1}^T
                       \mathbf{X}(t, k),~\mathrm{and}\\
  \sigma^2_{\mathbf{X}(\cdot, k)} &= \sum_{t=1}^T \left( \mathbf{X}(t, k) -
                            \bar{\mathbf{X}(\cdot, k)} \right)^2 
\end{align}
We can generalize this formula to compute time-varying correlations by
incorporating a \textit{kernel function} that takes a time $t$ as
input, and returns how much the observed data at each timepoint
$\tau \in \left[ -\infty, \infty \right]$ contributes to the estimated instantaneous
correlation at time $t$ (Fig.~\ref{fig:kernels}).
\begin{figure}
  \centering
  \includegraphics[width=\textwidth]{figs/kernels}
  \caption{\textbf{Examples of kernel functions.} Each panel
    displays per-timepoint weights at $t = 50$, evaluated for 100
    timepoints ($\tau \in \left[1, ..., 100\right]$).  \textbf{a. Uniform
      kernel.} The weights are timepoint-invariant; observations at
    all timepoints are weighted equally, and do not change as a
    function of $\tau$.  This is a special case of weight function that
    reduces dynamic correlations to static correlations.
    \textbf{b. Dirac $\delta$ kernel.} Only the observation at
    timepoint $t$ is given a non-zero weight (of 1).
    \textbf{c. Gaussian kernels.} Each kernel's weights fall off
    in time according to a Gaussian probability density function
    centered on time $t$.  Weights derived using several different
    example width parameters ($\sigma^2$) are displayed.
    \textbf{d. Laplace kernels.}  Each kernel's weights fall off
    in time according to a Laplace probability density function
    centered on time $t$.  Weights derived using several different
    example width parameters ($b$) are displayed.  \textbf{e. Mexican
      hat (Ricker wavelet) kernels.}  Each kernel's weights fall
    off in time according to a Ricker wavelet centered on time $t$.  This
    function highlights the \textit{contrasts} between local versus
    surrounding activity patterns in estimating dynamic
    correlations. Weights derived using several different example
    width parameters ($\sigma$) are displayed.}
  \label{fig:kernels}
\end{figure}

@@

\subsection*{Data}
We examined two types of data: synthetic data and human functional
neuroimaging data.  We constructed and leveraged the synthetic data to
evaluate our general approach.  Specifically, we tested how well
Equation~\ref{eqn:timecorr} could be used to recover known dynamic
correlations using different choices of kernel ($\kappa$;
Fig.~\ref{fig:kernels}), for each of several synthetic datasets that
exhibited different temporal properties.  We applied our approach to a
functional neuroimaging dataset to test the hypothesis that ongoing
cognitive processing is reflected in high-order dynamic correlations.
We used an across-participant classification test to estimate whether
dynamic correlations of different orders contain information about
which timepoint in a story participants were listening to.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section*{Discussion}
We tested the hypothesis that high-level cognition is supported by
high-order brain network dynamics.  We examined high-order network
dynamics in functional neuroimaging data collected during a story
listening experiment.  When participants listened to an auditory
recording of the story, participants exhibited similar high-order
brain network dynamics.  By contrast, when participants instead
listened to temporally scrambled recordings of the story, only
lower-order brain network dynamics were similar across participants.
Our results indicate that higher orders of network interactions
support higher-level aspects of cognitive processing
(Fig.\ref{fig:discussion}).

@@

@@

Relevant literature:
\begin{itemize}
\item \cite{SoloEtal19}: more complex concepts involve more complex network
dynamics

\item \cite{DemeEtal19}: human consciousness is supported by dynamic
complex patterns of brain signal coordination.  studied the brains of
healthy vs minimally conscious patients; conscious patients displayed
richer and more structured patterns of network interactions

\item \cite{MackEtal17}: richer task structure is supported by more complex (higher
rank) brain patterns in the prefrontal cortex

\item \cite{LuriEtal18}: review of time-varying connectivity

\item \cite{FongEtal19}: higher variability in dynamic functional connectivity
predicts worse attention.  also, dymamic functional connectivity
improves attention predictions over static functional connectivity.

\item \cite{ParkEtal18b}: dynamic causal modelling: compute causal interaction
dynamics in multivariate timeseries data.  important to consider
causation in interpreting the direction of information flow reflected
in the patterns we identified through our analyses.

\item \cite{PretEtal17}: another review of time varying connectivity

\item \cite{RoyEtal19}: memories are distributed across functionally connected
brain areas

\item \cite{FallEtal19}: the timescale of spontaneous resting state
  fMRI increases with structural connectivity strength (in mouse and
  human brains)

\item \cite{LiegEtal19}: dynamic resting state functional connectivity best captures
task-based phenotypes (processing speed, fluid intelligence), whereas
static and dynamic functional connectivity equally capture
self-reported measures like loneliness and life satisfaction.  suggest
that different timescales of connectivity dynamics reflect different
aspects of behavior.

\item \cite{ZhenEtal19}: characterize brain connectectomes at different geometric
scales

\item \cite{McInJirs19}: neural processes reflect system interactions that occur on
relatively low-dimensional manifolds, thereby constraining possible
functional configurations.  (motivation for dimensionality reduction
approach?)

\item \cite{TokeSomm19}: another approach to reducing dimensionality of network
interactions, using spectral clustering

\item \cite{ZouEtal19}: another review of complex network dynamics in timeseries
data

\item \cite{ShapEtal19}: state-based analyses of network dynamics using hidden
semi-Markov models.  could be a useful direction for extended our
work; for example, we could define kernel functions to only integrate
within an ``event.''  This could potentially improve resolution of
event-specific network dynamics.

\item \cite{ColeEtal19}: potential artifacts or spurious insights that affect
standard functional connectivity analyses (not sure this applies)

\item \cite{ThomEtal18}: simulations for benchmarking time-varying connectivity
methods (should we use this? seems useful! could potentially replace
synthetic data section)

\item \cite{ChanGlov10}: early citation arguing that dynamic
  functional connectivity might contain valuable signal that is
  different from static functional connectivity.

\item \cite{SizeEtal18}:

\item \cite{ReimEtal17}

\item \cite{GonzEtal19}

\item \cite{BetzEtal19}

\end{itemize}

>>>>

\section*{Introduction}
A central goal in cognitive neuroscience is to elucidate the
\textit{neural code}: the mapping between (a) mental states or
cognitive representations and (b) neural activity patterns. One means
of testing models of the neural code is to ask how accurately that
model is able to ``translate'' neural activity patterns into known (or
hypothesized) mental states or cognitive
representations~\citep[e.g.,][]{HaxbEtal01, NormEtal06, TongPrat12,
  MitcEtal08a, KamiTong05, NishEtal11, PereEtal18, HuthEtal12,
  HuthEtal16}.  Training decoding models on different types of neural
features (Fig.~\ref{fig:patterns}a) can also help to elucidate which
specific aspects of neural activity patterns are informative about
cognition-- and, by extension, which types of neural activity patterns
might comprise the neural code.  For example, prior work has used
region of interest analyses to estimate the anatomical locations of
specific neural representations~\citep[e.g.,][]{EtzeEtal09}, or to
compare the relative contributions to the neural code of multivariate
activity patterns versus patterns of dynamic correlations between
neural activity patterns~\citep[e.g.,][]{MannEtal18, FongEtal19}.  An
emerging theme in this literature is that cognition is mediated by
complex dynamic interactions between brain
structures~\citep{SporHone06, BassEtal06, Turk13, DemeEtal19,
  SoloEtal19, LuriEtal18, PretEtal17, ZouEtal19, MackEtal17}.

@@

Studies of the neural code to date have primarily focused on
univariate or multivariate neural patterns~\citep[for review
see][]{NormEtal06}, or (more recently) on patterns of dynamic
first-order correlations~\citep[i.e., interactions between pairs of
brain structures;][]{MannEtal18, FongEtal19, LuriEtal18, PretEtal17,
  ZouEtal19, DemeEtal19}.  We wondered what the
future of this line of work might hold.  For example, is the neural
code mediated by higher-order interactions between brain
structures~\citep[e.g., see][]{ReimEtal17}?
Second-order correlations reflect \textit{homologous} patterns of
correlation.  In other words, if the changing patterns of correlations
between two regions, $A$ and $B$, are similar to those between two
other regions, $C$ and $D$, this would be reflected in the
second-order correlations between ($A$--$B$) and ($C$--$D$).  In this
way, second-order correlations identify similarities and differences
between subgraphs of the brain's connectome.  Analogously, third-order
correlations reflect homologies between second-order correlations--
i.e., homologous patterns of homologous interactions between brain
regions.  More generally, higher-order correlations reflect homologies
between patterns of lower-order correlations.  We can then ask: which
``orders'' of interaction are most reflective of high-level cognitive
processes?

@@

To gain insights into the above aspects of the neural code, we
developed a computational framework for estimating dynamic high-order
correlations in timeseries data. This framework provides an important
advance, in that it enables us to examine patterns in higher-order
correlations that are computationally intractable to estimate via
conventional methods.  Given a multivariate timeseries, our framework
provides timepoint-by-timepoint estimates of the first-order
correlations, second-order correlations, and so on (up to tenth-order
correlations in this manuscript).  Our approach combines a
kernel-based method for computing dynamic correlations in timeseries
data with a dimensionality reduction step (Fig.~\ref{fig:patterns}b)
that projects the resulting dynamic correlations into a
low-dimensional space.  We explored two dimensionality reduction
approaches: principle components analysis~\citep[PCA;][]{Pear01},
which preserves an approximately invertable transformation back to the
original data~\citep[e.g., this follows related approaches taken
by][]{McInJirs19, TokeSomm19, GonzEtal19}; and a second non-invertible
algorithm that explored patterns in eigenvector
centrality~\citep{Land95}.  This latter approach characterizes
correlations between each feature dimension's relative
\textit{position} in the network in favor of the specific activity
histories of different features~\citep[also see][]{BetzEtal19,
  SizeEtal18, ReimEtal17}.

@@

\subsection*{Kernel-based approach for computing dynamic correlations}
Given a matrix of observations, we can compute the (static)
Pearson's correlation between any pair of columns, $\mathbf{X}(\cdot, i)$ and
$\mathbf{X}(\cdot, j)$ using~\citep{Pear01}:
\begin{align}
  \mathrm{corr}(\mathbf{X}(\cdot, i), \mathbf{X}(\cdot, j)) &=
                                                              \frac{\sum_{t=1}^T
                                                              \left(\mathbf{X}(t,
                                                              i)
                                                              -
                                                              \bar{\mathbf{X}(\cdot,
                                                              i)}\right)
                                                              \left(\mathbf{X}(t,
                                                              j)
                                                              -
                                                              \bar{\mathbf{X}(\cdot, j)}\right)}{\sqrt{\sum_{t=1}^T
                                                              \sigma^2_{\mathbf{X}(\cdot, i)} 
                                                              \sigma^2_{\mathbf{X}(\cdot, j)}}},~\mathrm{where}\\\label{eqn:corr}
  \bar{\mathbf{X}(\cdot, k)} &= \sum_{t=1}^T
                       \mathbf{X}(t, k),~\mathrm{and}\\
  \sigma^2_{\mathbf{X}(\cdot, k)} &= \sum_{t=1}^T \left( \mathbf{X}(t, k) -
                            \bar{\mathbf{X}(\cdot, k)} \right)^2 
\end{align}
We can generalize this formula to compute time-varying correlations by
incorporating a \textit{kernel function} that takes a time $t$ as
input, and returns how much the observed data at each timepoint
$\tau \in \left[ -\infty, \infty \right]$ contributes to the estimated instantaneous
correlation at time $t$~\citep[Fig.~\ref{fig:kernels}; also see][for a
similar approach]{AlleEtal12b}.
\begin{figure}
  \centering
  \includegraphics[width=\textwidth]{figs/kernels}
  \caption{\textbf{Examples of kernel functions.} Each panel
    displays per-timepoint weights at $t = 50$, evaluated for 100
    timepoints ($\tau \in \left[1, ..., 100\right]$).  \textbf{a. Uniform
      kernel.} The weights are timepoint-invariant; observations at
    all timepoints are weighted equally, and do not change as a
    function of $\tau$.  This is a special case of weight function that
    reduces dynamic correlations to static correlations.
    \textbf{b. Dirac $\delta$ kernel.} Only the observation at
    timepoint $t$ is given a non-zero weight (of 1).
    \textbf{c. Gaussian kernels.} Each kernel's weights fall off
    in time according to a Gaussian probability density function
    centered on time $t$.  Weights derived using several different
    example width parameters ($\sigma^2$) are displayed.
    \textbf{d. Laplace kernels.}  Each kernel's weights fall off
    in time according to a Laplace probability density function
    centered on time $t$.  Weights derived using several different
    example width parameters ($b$) are displayed.  \textbf{e. Mexican
      hat (Ricker wavelet) kernels.}  Each kernel's weights fall
    off in time according to a Ricker wavelet centered on time $t$.  This
    function highlights the \textit{contrasts} between local versus
    surrounding activity patterns in estimating dynamic
    correlations. Weights derived using several different example
    width parameters ($\sigma$) are displayed.}
  \label{fig:kernels}
\end{figure}

@@

\subsection*{Data}
We examined two types of data: synthetic data and human functional
neuroimaging data.  We constructed and leveraged the synthetic data to
evaluate our general approach~\citep[for a related approach,
see][]{ThomEtal18}.  Specifically, we tested how well
Equation~\ref{eqn:timecorr} could be used to recover known dynamic
correlations using different choices of kernel ($\kappa$;
Fig.~\ref{fig:kernels}), for each of several synthetic datasets that
exhibited different temporal properties.  We applied our approach to a
functional neuroimaging dataset to test the hypothesis that ongoing
cognitive processing is reflected in high-order dynamic correlations.
We used an across-participant classification test to estimate whether
dynamic correlations of different orders contain information about
which timepoint in a story participants were listening to.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section*{Discussion}
We tested the hypothesis that high-level cognition is supported by
high-order brain network dynamics~\citep[e.g., see][]{SoloEtal19,
  ReimEtal17}.  We examined high-order network dynamics in functional
neuroimaging data collected during a story listening experiment.  When
participants listened to an auditory recording of the story,
participants exhibited similar high-order brain network dynamics.  By
contrast, when participants instead listened to temporally scrambled
recordings of the story, only lower-order brain network dynamics were
similar across participants.  Our results indicate that higher orders
of network interactions support higher-level aspects of cognitive
processing (Fig.\ref{fig:discussion}).

@@

The notion that cognition is reflected in (and possibly mediated by)
patterns of network dynamics has been suggested by or proposed in
myriad empirical studies and reviews~\citep[e.g., ][]{DemeEtal19,
  Turk13, LuriEtal18, FongEtal19, ParkEtal18b, PretEtal17,
  RoyEtal19, LiegEtal19, ZouEtal19, ChanGlov10, GonzEtal19}.  Our
study extends this line of work by finding cognitively relevant
higher-order network dynamics that reflect ongoing cognition.  Our
findings complement other work that uses graph theory and topology to
characterize how brain networks reconifigure during
cognition~\citep[e.g., ][]{BassEtal06, ZhenEtal19, McInJirs19,
  TokeSomm19, SizeEtal18, ReimEtal17, BetzEtal19}.

An open question not addressed by our study pertains to how different
structures integrate incoming information with different time
constants.  For example, one line of work suggests that the cortical
surface comprises a structured map such that nearby brain structures
process incoming information at similar timescales.  Low-level sensory
areas integrate information relatively quickly, whereas higher-level
regions integrate information relatively slowly~\citep{BaldEtal17,
  HassEtal08, HassEtal15, HoneEtal12a, LernEtal11, LernEtal14}.  Other
related work in human and mouse brains indicates that the temporal
response profile of a given brain structure may relate to how strongly
connected that structure is with other brain
areas~\citep{FallEtal19}.  Further study is needed to understand the
role of temporal integration at different scales of network
interaction, and across different anatomical structures.

Another potential limitation of our approach relates to recent work
suggesting that the brain undergoes rapid state changes, for example
across event boundaries~\citep[e.g., ][]{BaldEtal17}.
\cite{ShapEtal19} used hidden semi-Markov models to estimate
state-specific network dynamics.  Our general approach might be
extended by considering putative state transitions. For example,
rather than weighting all timepoints using a similar kernel
(Eqn.~\ref{eqn:timecorr}), the kernel function could adapt on a
timepoint-by-timepoint basis such that only timepoints determined to
be in the same ``state'' were given non-zero weight.

@@

\subsection*{Concluding remarks}
The complex hierarchy of dynamic interactions that underlie our
thoughts is perhaps the greatest mystery in modern science.  Methods
for characterizing the dynamics of high-order correlations in neural
data provides a window into the neural basis of cognition.
\textbf{Need a good closing sentence.}
